\vspace{-.5cm} Lights: a generalized joint model for high-dimensional multivariate longitudinal data and censored durations \vspace{.5cm}

### 2022-04-12 slides/JDS2022/short paper/paper.tex
add JDS pres folder

\section{Introduction}

With the increasing expectations to know their customers from account opening throughout the duration of the business relationship, web companies have the luxury of building elaborate systems to help them keep everything on track. The amount of recorded data per client is often tremendous and growing through time. There is no tool today to take into account simultaneously a huge number of longitudinal signals in a high-dimensional context to perform real-time churn (or satisfaction) risk prediction.
Similarly, in many clinical studies, it has become increasingly common to record the values of longitudinal features (e.g., biomarkers) until the occurrence of an event of interest for a subject. The ``joint modeling'' approaches, namely modeling the longitudinal and survival outcomes through a joint likelihood model rather than separately, has received considerable attention during the past two decades~\citep{tsiatis2004joint}. Numerical studies suggest that these approaches are among the most satisfactory to combine information~\citep{yu2004joint}. They have the advantage of making more efficient use of the data since information about survival also goes into modeling the longitudinal features. In addition, they produce unbiased estimates and do not rely on approximations for incorporating conplex longitudinal trajectories. Most developments have either focused on shared random-effect models (SREMs)~\citep{wulfsohn1997joint}, in which characteristics of the longitudinal processes (for instance functions of the random effects) are included as features in the survival model ; or on joint latent class models (JLCMs)~\citep{vermunt2003latent}, in which the population is considered as heterogeneous, with the assumption that there exist homogeneous latent classes sharing the same marker trajectories and the same prognostic.

\paragraph{The high-dimensional longitudinal data context.}
With the exploding number of daily internet users, or with the development of electronic health records in a medical context, high-dimensional settings are becoming increasingly frequent in various contexts where the number of available features to consider as potential risk factors is tremendous.
Moreover, with an increased focus on personalised medicine, the need to implement multivariate models that account for a large number of longitudinal outcomes is critical. Despite this, joint models have predominantly focused on univariate data, with attempts to fit multiple univariate joint models separately~\citep{wang2012joint}, which is inefficient~\citep{lin2002maximum}. 
Despite many multivariate models being presented in full generality, questions arrising from the high-dimensional context -- e.g., computational power, limits in numerical estimation, or sample size -- are never considered in analyses (to the best of our knowledge), and the number of longitudinal outcomes considered in numerical studies are often very low (see~\citet{hickey2016joint} for a complete review). For instance,~\citet{jaffa2014joint} only considers 3 longitudinal outcomes in the simulation study while mentioning a ``high-dimensional multivariate longitudinal data'' context.

\paragraph{General framework.}

The setting of this paper is such that we want to incorporate high-dimensional time-dependent (longitudinal) features measured with error in a survival model. Let us consider the usual survival analysis framework.
Following~\citet{andersen2012statistical}, let non-negative random variables $T^\star$ and $C$ stand for the times of the event of interest and censoring times respectively. The event of interest could be for instance the time when a client stops using a company's product or service in a churn prediction setting~ ; or survival time, re-hospitalization, relapse or disease progression in a medical context.
We then denote $T$ the right-censored time and $\Delta$ the censoring indicator, defined as 
\begin{equation*}
T = T^\star \wedge C \quad \text{and} \quad \Delta = \ind{\{T^\star \leq C\}}
\end{equation*}
respectively, where $a \wedge b$ denotes the minimum between two numbers $a$ and $b$, and $\ind{\{\cdot\}}$ the indicator function taking the value $1$ if the condition in $\{\cdot\}$ is satisfied and $0$ otherwise.

Let $X$ denotes the $p$-dimensional vector of time-independent features (e.g., patients characteristics, therapeutic strategy, or omics features recorded at the begining of a medical study), and let  $Y(t) = \big(Y^1(t), \ldots, Y^L(t) \big)^\top \in \R^L$ denote the value of the $L$-dimensional longitudinal outcome at time point $t \geq 0$, with $L \in \N_+$.

\paragraph{Heterogeneity of the population.}

An assumption of heterogeneity within the subject population is frequently relevant in medical research where several differing profiles of subjects are expected~\citep{bussy2019c}. To take account of this, we introduce a latent variable $G \in \{0, \ldots, K-1\}$ modeling the $K \geq 1$ classes of different risk, which is a classical modeling assumption in JLCMs~\citep{lin2002latent, proust2014joint}. Let us denote
\begin{equation}
  \label{eq:pi}
  \pi_{\xi_k}(x) = \P[G=k|X=x]
\end{equation}
the latent class membership probability given time-independent features $x \in \R^p$, and consider a softmax link function given by
\begin{equation}
  \label{eq:pi-softmax} 
  \pi_{\xi_k}(x) = \dfrac{e^{x^\top\xi_k}}{\sum_{k=0}^{K-1}e^{x^\top\xi_k}}
\end{equation}
where $\xi_k \in \R^p$ denotes a vector of coefficients that quantifies the impact of each time-independent features on the probability that a subject belongs to the $k$-th latent class, with $\xi_0 = \mathbf{0}_p$ for overparameterization purpose, where $\mathbf{0}_p$ stands for the vector of $\R^p$ having all coordinates equal to zero. The intercept term is here omitted without loss of generality. 
From now on, all computations are done conditionally on features $x$.

\paragraph{Main contribution.}

In this paper, we propose a method called \textit{lights} (generaLized joInt hiGH-dimensional longiTudinal Survival) which is inspired by both JLCMs and SREMs, since we also include features extracted from the longitudinal processes as potential risk factor in the survival model, which is a class-specific Cox model~\citep{Cox1972JRSS} with high-dimensional shared associations.
To allow flexibility in modeling the dependency between the longitudinal features and the event time, we use appropriate penalties : elastic net~\citep{zou2005regularization}
for time-independent feature selection in the latent class membership, and sparse group lasso~\citep{simon2013sparse} in the survival model : this penalty acts like the lasso at the trajectory level, namely an entire trajectory may drop out of the model, and yields sparsity for a given trajectory.
Inference is achieved using an efficient and novel proximal Quasi-Newton Expectation Maximization algorithm (prox-QNEM).
Posterior estimates of the latent class membership probabilities given the longitudinal process and the fact that the subject is still alive at prediction time -- in a so called ``real-time prediction paradigm'' -- are then used as discriminative marker rule for risk prediction in the cross-validation procedure for selecting the best regularization hyper-parameters.
Hence, the method provides interpretations of the high-dimensional longitudinal features, thus offering a powerful tool for real-time decision support in a customer's satisfaction monitoring context for instance, or else for clinical decision making in patient monitoring.

\paragraph{Organization of the paper.}

A precise description of the model is given in Section~\ref{sec:Method}.
Section~\ref{sec:inference} focuses on a regularized version of the
model to exploit dimension reduction and prevent overfitting. Inference is
presented under this framework, as well as the developed algorithm.
Section~\ref{sec:Performance evaluation} introduces the C-index metric, as well as a novel evaluation strategy to assess diagnostic prediction performances while mimicking a real-time use of the model in clinical care, and finally the considered competing methods.
Section~\ref{sec:simulation study} presents the simulation procedure used to evaluate the performance of our method in a high-dimensional context and compares it with state-of-the-art ones. In Section~\ref{sec:application}, we apply our method to a publicly available dataset. Finally, we discuss the obtained results in Section~\ref{sec:conclusion}.

\paragraph{Notations.}

Throughout the paper, for every $q > 0,$ we denote by $\norm{v}_q$ the usual $\ell_q$-quasi norm of a vector $v \in \R^m,$ namely $\norm{v}_q =(\sum_{k=1}^m|v_k|^q)^{1/q}$. We also denote $\norm{v}_0 = |\{k : v_k \neq 0\}|$, where $|A|$ stands for the cardinality of a finite set $A$. $\lfloor a \rfloor$ denotes the largest integer less than or equal to $a \in \R$.
For $u, v \in \R^m$, we denote by $u \odot v$ the Hadamard product $u\odot v =(u_1v_1, \ldots, u_mv_m)^\top.$ 
For a squared matrix $M$, $\text{vech}(M)$ stacks columns of $M$ one under another in a single vector, starting each column at its diagonal element. We write $I_m$ for the identity matrix of $\R^{m \times m}$.
Finally, we write, for short, $\mathbf{1}_m$  (resp. $\mathbf{0}_m$) for the vector of $\R^m$ having all coordinates equal to one (resp. zero).

\section{Method}
\label{sec:Method}

In this section, we descibe the longitudinal and time-to-event submodels, as well as the required hypothesis in order to write a likelihood and draw inference for the lights model.

\subsection{Class-specific marker trajectories}

We suppose a class-specific marker trajectory and a generalized linear mixed model for each longitudinal marker given latent class $G$, so that for the $l$-th outcome at time $t \geq 0$ one has
\begin{equation}
  \label{eq:link-function}
  h_l\big(\E[Y^l(t)|b^l, G=k]\big) = m_k^l(t)
 \end{equation}
where $h_l$ denotes a known one-to-one link function, and $m_k^l$ the linear predictor such that
\[ m_k^l(t) = u^l(t)^\top\beta_k^l + v^l(t)^\top b^l \]
where $u^l(t) \in \R^{q_l}$ is a row vector of (possibly) time-varying features with corresponding unknown fixed effect parameters $\beta_k^l$, and $v^l(t) \in \R^{r_l}$ is a row vector of (possibly) time-varying features with $r_l \leq q_l$.
A suitable distributional assumption for the random effects component is a zero-mean multivariate normal distribution~\citep{hickey2016joint}, that is
\begin{equation}
   \label{eq:random-effect-law}
   b^l \sim \cN(0, D_{ll})
 \end{equation} 
with $D_{ll} \in \R^{r_l \times r_l}$ the unstructured variance-covariance matrix. 
To account for dependence between the different longitudinal outcome types, we let $\text{Cov}[b^l,b^{l'}] = D_{ll'}$ for $l \ne l'$ and we denote
\[ D = 
\begin{bmatrix}
  D_{11} & \cdots & D_{1L}\\
  \vdots &  \ddots & \vdots \\
  D_{1L}^\top & \cdots & D_{LL}
\end{bmatrix}
\]
the global variance-covariance matrix.
In the sequel of the paper, our aim is to associate the true and unobserved value $m_k^l(t)$ of the $l$-th longitudinal outcome at time $t$ with the event outcome $T^\star$.

\subsection{Class-specific risk of event}

To quantify the effect of the longitudinal outcomes on the risk for an event, we use a Cox~\citep{Cox1972JRSS} relative risk model of the form
\begin{equation}
  \label{eq:intensity-model}
  \lambda(t|\cY(t), G = k) = \lambda_0(t) \exp \Big\{\sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l} \Psi_a^l(t) \Big\},
\end{equation}
where $\lambda_0(\cdot)$ is an unspecified baseline hazard function and given that $G = k$, we denote 
\[\cY^l(t) = \{Y^l(u), 0 \leq u < t\} \quad \text{and} \quad \cY(t) = \bigcup_{l=1}^L \cY^l(t)\] 
the history of the true longitudinal process up to time $t$.
For each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known functionals $\Psi_a^l$ extracted from $\cY^l(t)$ through a given representation mapping, and $\gamma_{k,a}^l \in \R$ the corresponding joint representation parameters.

Let us finally denote
\[\gamma_k= (\gamma_{k,1}^1, \ldots, \gamma_{k,\cA}^1, \ldots, \gamma_{k,1}^L, \ldots, \gamma_{k,\cA}^L )^\top \in \R^{L\cA}, \] so that
\begin{equation}
  \label{eq:def-asso-features}
  \lambda(t|\cY(t), G = k) = \lambda_0(t) \exp \big\{\gamma_k^\top \Psi(t) \big\},
\end{equation}
where $\Psi(t) \in \R^{L\cA}$ denotes the representation features vector.

\subsection{Generalization of SREMs and JLCMs}

Our model is clearly of JLCMs type, and can be viewed as a generalization of SREMs type in a much more flexible way. Table~\ref{table:shared_associations} gives common parameterizations for shared associations that are only considered separately ($\cA=1$) in SREMs.
\begin{table}[htb]
\centering
\begin{tabular}{cccc}
\toprule
Description & $\Psi_a^l(t, \beta_k^l, b^l)$  & Reference \\
\midrule
Linear predictor & $m_k^l(t)$ & \citet{chi2006joint} \\ [.15cm]
Random effects & $b^l$  & \citet{hatfield2011joint} \\
Time-dependent slope & $\dfrac{\dd}{\dd t} m_k^l(t)$ & \citet{rizopoulos2011bayesian} \\ [.3cm]
Cumulative effect & $\int_0^t m_k^l(s) \dd s$ & \citet{andrinopoulou2017combined} \\ [.1cm]
 \bottomrule
\end{tabular}
\caption[]{Examples of classical shared associations in SREMs.}
\label{table:shared_associations}
\end{table}

Our proposition differs from standard SREMs in two respects: $(i)$ representation vector $\Psi(t)$ does not depend on the modeling assumptions in the longitudinal submodel and $(ii)$ it allows to choose multiple high-dimensional representation mappings that characterize time series and concatenate them, since we perform features selection through the regularization strategy described in Section~\ref{sec:penalized-obj}. Point $(i)$ is key since for instance the $\beta_k^l$ dependence in $\Psi_a^l$ would lead to complicated updates, while the $b^l$ dependence would lead to untracktable integrals that are common in SREMs and require approximation methods such as Monte Carlo~\citep{hickey2018joinerml}, being computationally intensive and not scalable in a high-dimensional context. For point $(ii)$ in practice, we use the \texttt{Python} library \texttt{tsfresh}~\citep{christ2018time} and include many extracted features, such as absolute energy of the time series, statistics on autocorrelation, or Fourier and wavelet basis projections, to name but a few.

\subsection{Graphical model}
Figure~\ref{fig:graphical-model} below gives the graphical representation for the lights model to get a beter view of the dependence structure for the model.
\begin{figure}[!htb]
\centering
\resizebox{.9\textwidth}{!}{
\begin{tikzpicture}
\tikzstyle{main}=[circle, minimum size = 18mm, ultra thick, draw =black!80, node distance = 13mm,text centered]
\tikzstyle{var}=[rectangle, minimum size = 16mm, ultra thick, draw =black!80, node distance = 13mm,text centered]
\tikzstyle{connect}=[-latex, thick]
\tikzstyle{box}=[rectangle, draw=black!100]
  \node[var, fill = white!100] (beta) [] {\Large$\bm{\beta_k^l}$};
  \node[main, fill = white!100] (mu) [below=4cm of beta] {$\bm{M_{ik}^l(t)}$};
  \node[main, fill = black!30] (Y) [below=of mu] {\Large$\bm{Y_i^l(t)}$};
  \node[main, fill = white!100] (b) [left=of mu] {\Large$\bm{b_i^l}$};
  \node[var, fill = white!100] (phi) [left=of Y] {\Large$\bm{\phi_l}$};
  \node[var, fill = white!100] (Sigma) [above=of b] {\Large$\bm{D}$};
  \node[var, fill = white!100] (xi) [right=4.5cm of beta] {\Large$\bm{\xi_k}$};
  \node[var, fill = white!100] (gamma) [right=5cm of xi] {\Large$\bm{\gamma_k}$};
  \node[main, fill = white!100] (G) [below=4cm of xi] {\Large$\bm{G_i}$};
  \node[main, fill = white!100] (W) [right=8.5cm of Y] {\Large$\bm{\Psi_i(t)}$};
  \node[main, fill = white!100] (lambda) [below=4cm of gamma] {\Large$\bm{\lambda_i(t)}$};
  \node[main, fill = black!30] (T) [below=1.3cm of lambda] {\Large$\bm{T_i}$};
  \node[main, fill = black!30] (X) at (4, -3.3) {\Large$\bm{X_i}$};

  \path [->,draw,ultra thick, blue]
        (G) edge (mu)
        (Y) edge (W);
  \path [->,draw,ultra thick]
        (beta) edge (mu)
        (b) edge (mu)
        (mu) edge (Y)
        (Sigma) edge (b)
        (phi) edge (Y)
        (xi) edge (G)
        (X) edge (G)
        (gamma) edge (lambda)
        (W) edge (lambda)
        (G) edge (lambda)
        (lambda) edge (T)
        ;

  \node[rectangle, inner sep=8mm,draw=black!130, ultra thick, fit= (beta) (xi) (gamma),label={[anchor=west, yshift=-0.4cm, xshift=-8cm]{\Large$\bm{k = 1, \ldots,K}$}}] at (6.8,0.4) {};
  \node[rectangle, inner sep=4mm,draw=black!100, ultra thick, fit= (beta) (mu) (b) (Y) (Sigma) (phi)] at (-1.6,-4.5) {};
  \node[rectangle, inner sep=4mm, fit= (beta) (mu) (b) (Y) (Sigma) (phi),ultra thick, label={[anchor=west, yshift=-0.5cm, xshift=-2.8cm]{\Large$\bm{l = 1, \ldots,L}$}}] {};

  \node[rectangle,inner sep=6mm, draw=black!100, label=center:{\color{white}\Large{\textbf{Longitudinal}}}, minimum width=5cm, minimum height = 1cm, fill = black!50] at (-1.5,2.8) {};

  \node[rectangle,inner sep=6mm, draw=black!100,  label=center:{\color{white}\Large{\textbf{Time-independent}}}, minimum width=5cm, minimum height = 1cm, fill = black!50] at (5.2,2.8) {};

  \node[rectangle,inner sep=6mm, draw=black!100,  label=center:{\color{white}\Large{\textbf{Time-to-event}}}, minimum width=5cm, minimum height = 1cm, fill = black!50] at (12,2.8) {};

\end{tikzpicture}}
\caption{Graphical representation of the lights model. Square boxes denote observed data while circles denote unobserved (including random) terms. One can then observe the two links in blue between the submodels that gives rise to the joint model.}
\label{fig:graphical-model}
\end{figure}

\subsection{Likelihood}

Consider an independent and identically distributed (i.i.d.) cohort of $n$ subjects
\[ \cD_n = \big\{ (x_1, y_1^1, \ldots, y_1^L, t_1, \delta_1), \ldots, (x_n, y_n^1, \ldots, y_n^L, t_n, \delta_n) \big\} \]
where for each subject $i=1, \ldots, n$, process $Y_i^l$ is measured $n_i^l$ times at $t_{i1}^l, \ldots, t_{in_i^l}^l$ (which can differ between subjects and outcomes) with $t_{ij}^l \leq t_{ij+1}^l$ for all $j=1, \ldots, n_i^l-1$ and such that
\[ y_i^l=(y_{i1}^l, \ldots, y_{in_i^l}^l)^\top \in \R^{n_i^l} \quad \text{ with } \quad y_{ij}^l=Y_i^l(t_{ij}^l) \]
for all $l=1, \ldots, L$. 
For the $i$-th subject, let us denote 
\[
\left\{
    \begin{array}{ll}
        y_i &= ({y_i^1}^\top \cdots {y_i^L}^\top)^\top \in \R^{n_i},\\
        b_i &= ({b_i^1}^\top \cdots {b_i^L}^\top)^\top \in \R^r,
    \end{array}
\right.
\]
with $n_i = \sum_{l=1}^L n_i^l$ and $r = \sum_{l=1}^L r_l$ the total number of longitudinal measurements (for subject $i$) and the total dimension of the random effects respectively, as well as the following design matrices
\[ U_i = 
\begin{bmatrix}
  U_{i1} & \cdots & 0\\
  \vdots &  \ddots & \vdots \\
  0 & \cdots & U_{iL}
\end{bmatrix} 
\in \R^{n_i \times q}
\qquad \text{and} \qquad
V_i = 
\begin{bmatrix}
  V_{i1} & \cdots & 0\\
  \vdots &  \ddots & \vdots \\
  0 & \cdots & V_{iL}
\end{bmatrix}
\in \R^{n_i \times r}
\]
with $q = \sum_{l=1}^L q_l$ and where for all $l=1, \ldots, L$, one writes
\[
\left\{
    \begin{array}{ll}
        U_{il} &= \big(u_i^l(t_{i1}^l)^\top \cdots u_i^l(t_{in_i^l}^l)^\top\big)^\top \in \R^{n_i^l \times q_l},\\
        V_{il} &= \big(v_i^l(t_{i1}^l)^\top \cdots v_i^l(t_{in_i^l}^l)^\top\big)^\top \in \R^{n_i^l \times r_l}.
    \end{array}
\right.
\]
From now on, all computations are done conditionally on the design matrices $(U_i)_{i=1, \dots, n}$ and $(V_i)_{i=1, \dots, n}$.
Denoting $(t_1^u, \ldots, t_J^u)$ the $J \in \N_+$ unique failure times, we then extract all representation vectors $\big(\Psi_i(t_j^u)\big)_{j=1,\ldots,J}$ for all subjects $i=1, \dots, n$ before inference and from now on, all computations are also done conditionally on them.
Now, let us make two first hypothesis.
\begin{assumption}
\label{assumption1}
Suppose that the random effects are independent of the latent class membership, and that the latter remain independent conditional on the observed data (namely $T$, $\Delta$ and $Y$).
\end{assumption}
Assumption~\ref{assumption1} states that subject-and-longitudinal outcome specific random effects $b^l$ do not depend on the latent class membership, which is not a strong modeling assumption.
\begin{assumption}
\label{indep-hyp-2}
Suppose that for a given subject $i$ and conditional on the random effects and the latent class membership, all $Y_i^l$ are independent for $l=1, \dots, L$.
\end{assumption}
Assumption~\ref{indep-hyp-2} is similar to standard modeling hypothesis, see for instance~\citet{tsiatis2004joint}.
For all $k = 0, \ldots, K-1$, we denote 
\[ \beta_k = ({\beta_k^1}^\top \cdots {\beta_k^L}^\top)^\top \in \R^q \] 
and 
\[M_{ik} = U_i\beta_k + V_ib_i \in \R^{n_i}. \]

Given~\eqref{eq:link-function} and Assumption~\ref{indep-hyp-2}, each $y_i^l$ is assumed to be from a one-parameter exponential family with respect to a reference measure which is either the Lebesgue measure (e.g., in the Gaussian case) or the counting measure (e.g., in the logistic case).
The conditional distribution of $y_i|b_i, G_i=k$ is then assumed to be from a distribution with a density of the form
\begin{equation}
	\label{eq:exp-family-density}
	f(y_i|b_i, G_i=k) = \exp \big\{(y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \big\},
\end{equation}
with 
\begin{equation}
  \label{eq:def-Phi_i} 
  \Phi_i = (\phi_1^{-1} {\textbf{1}_{n_i^1}}^\top \cdots \phi_L^{-1} {\textbf{1}_{n_i^L}}^\top)^\top \in \R^{n_i}
\end{equation} 
and $\phi = (\phi_1, \ldots, \phi_L)^\top \in \R^L$.
The density described in~\eqref{eq:exp-family-density} encompasses several distributions, see Table~\ref{table:glm}. The functions $c_\phi(\cdot)$ and $d_\phi(\cdot)$ are known as well as the dispersion parameter $\phi_l$, while parameter $\beta_k$ has to be estimated. Note that $d_\phi(\cdot)$ is related to the normalizing constant.
\begin{table}[htb]
\centering
\resizebox{\textwidth}{!}{%
\begin{tabular}{cccccc}
\toprule
 Model & Support & Use cases & $\phi_l$ & $h_l(\cdot)$ in~\eqref{eq:link-function} & $c_\phi(\cdot)$ \\
 \midrule
 Gaussian & $\R$ & Continuous response data & $\sigma_l^2$ & $z \mapsto z$ & $z \mapsto z^2 / 2 \sigma_l^2$ \\ 
 Categorical & $\{0, 1\}^C$ & Outcome with $C$ modalities & 1 & $z \mapsto \log\Big(\dfrac{z}{1 - z} \Big)$ & $z \mapsto \log(1 + e^z)$ \\
 Poisson & $\N$ & Count of occurrences & 1 &  $z \mapsto \log(z)$ & $z \mapsto e^z$ \\
 \bottomrule
\end{tabular}}
\caption[]{Example of standard distributions that fit in the considered setting, given in the univariate case for simplicity of the notations.}
\label{table:glm}
\end{table}

Let us denote 
\[ \theta = \big(\xi_0^\top \cdots \xi_{K-1}^\top, \beta_0^\top \cdots \beta_{K-1}^\top, \phi^\top, \text{vech}(D), \lambda_0^\top, \gamma_0^\top \cdots \gamma_{K-1}^\top\big)^\top \in \R^\vartheta \]
the collection of the $\vartheta \in \N_+$ unknown parameters to estimate. Note that we include $\lambda_0$ in $\theta$. Indeed, we will estimate the baseline using nonparametric maximum likelihood, as a function taking mass at each failure time. So in $\theta$, $\lambda_0$ represents the vector of these baseline hazard quantities, with a dimension equal to $J$.
To write the log-likelihood $\ell_n(\theta)$ (rescaled by $n^{-1}$)  for samples in $\cD_n$, corresponding to the joint distribution of the time-to-event and longitudinal outcomes, let us make the following hypothesis.
\begin{assumption}
\label{indep-hyp-3}
Assume that
\begin{equation}
  \label{eq:ind-hyp}
  f_\theta(t_i, \delta_i, y_i | b_i, G_i = k) = f_\theta(t_i, \delta_i| G_i = k) f_\theta(y_i | b_i, G_i = k)
\end{equation}
for all $i=1, \ldots, n$.
\end{assumption}
Assumption~\ref{indep-hyp-3} is a generalization of classical hypothesis used in SREMs and JLCMs~\citep{hickey2016joint}. Remember that the densities are also conditionned by all the representation features $\big(\Psi_i(t_j^u)\big)_{j=1,\ldots,J}$. Hence Assumption~\ref{indep-hyp-3} says that the random effects vector $b_i$, the latent class membership and the representation features account for the association between the longitudinal and event outcomes, which makes sense given Figure~\ref{fig:graphical-model}.
Then, one has
\begin{equation}
	\label{eq:log-lik}
	\ell_n(\theta) = \ell_n(\theta ; \cD_n) = n^{-1} \sum_{i=1}^n \log \sum_{k=0}^{K-1} \pi_{\xi_k}(x_i) f_\theta(t_i, \delta_i| G_i = k) f_\theta(y_i | G_i = k),
\end{equation}
where 
\[f_\theta(t_i, \delta_i| G_i = k) = \big[\lambda(t_i|\cY_i(t_i), G_i = k)\big]^{\delta_i} \exp \Big\{-\int_0^{t_i} \lambda(s|\cY_i(s), G_i = k) \dd s \Big\}. \]
As expected, one has a JLCMs like log-likelihood.

\paragraph{Specification of the design matrices.}

In many practical applications, subjects show highly nonlinear longitudinal trajectories. We consider a flexible representations for $u^l(t)$ using a high-dimensional vector of time monomials, namely 
\begin{equation}
  \label{eq:alpha-def}
  u^l(t) = (1, t, t^2, \ldots, t^\alpha)^\top
\end{equation}
with $\alpha \in \N_+$.
The idea here is to let the practitioner choose a suitable polynomial order $\alpha$ for the representation -- that could also be tuned automatically using a model selection procedure and AIC~\citep{akaike1974new} for instance.
We then let \[v^l(t) = (1, t)^\top\] so that each trajectory of each subject gets an affine random effect. Hence with this choice in practice, one has $q_l=\alpha + 1$ and $r_l=2$ for all $l=1, \ldots, L$.

\section{Inference}
\label{sec:inference}

In this section, we describe the procedure for estimating the parameters of the lights model. Let us first present the considered penalized objective, and then focus on the algorithm proposed for inference.

\subsection{Penalized objective}
\label{sec:penalized-obj}

In order to avoid overfitting and improve the prediction power of our method, we propose to minimize the penalized objective
\begin{equation}
  \label{eq:pen-log-lik}
	\ell_n^\text{pen}(\theta) = - \ell_n(\theta) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}
\end{equation}
where for all $k=0, \ldots, K-1$, we add an elastic net regularization~\citep{zou2005regularization} of the vector $\xi_k$ and a sparse group lasso regularization~\citep{simon2013sparse} of the vector $\gamma_k$, for tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k})^\top \in \R_+^2$. Here, for fixed $(\eta, \tilde{\eta}) \in [0, 1]^2$, one has
\[ \norm{z}_{\text{en}, \eta} = (1-\eta)\norm{z}_1 + \dfrac\eta2 \norm{z}_2^2 \]
for the elastic net penalty of any vector $z$, that is a linear combination of the lasso ($\ell_1$) and ridge (squared $\ell_2$) penalties, and
\[ \norm{z}_{\text{sg} l_1, \tilde{\eta}} = (1-\tilde{\eta})\norm{z}_1 + \tilde{\eta} \sum_{l=1}^L\norm{z}_{2,l} \] for the sparse group lasso penalty of any vector $z$, with $\norm{z}_{2,l}=\norm{z^l}_2$, and for which we do not have to account for group sizes since they all have the same one. 
Hence, the resulting optimization problem is written
\begin{equation}
  \label{eq:optim-pb}
   \hat \theta \in \argmin_{\theta \in \R^\vartheta} \ell_n^\text{pen}(\theta).
 \end{equation}

One advantage of the considered regularization strategy is its ability to perform feature selection and pinpoint the most important (time-dependent and time-independent) features relatively to the prediction objective. The support of $\hat \xi_k$ informs on the time-independent features involved in the $k$-th latent class membership. The ridge part allows to handle potential correlation between time-independent features.
On the other hand, for the sparse group lasso penalty, a group $l$ corresponds to a trajectory, that is one longitudinal outcome. So if $\hat \gamma_k^l$ is entirely zeroed (thanks to the group lasso part), it means that the $l$-th longitudinal process is discarded by the model in terms of risk effect for the $k$-th latent class.
Then, the sparse part of the penalty allows to perform representation features selection for each trajectory : for $\hat \gamma_k^l$ that are not entirely zeroed, its support informs on the representation features involved in the $k$-th latent class risk of event for the $l$-th longitudinal outcome.
Note that in practice, the intercept terms are not regularized.

\subsection{A Proximal Quasi-Newton EM}
\label{sec:prox-QNEM}

In order to derive an algorithm for this objective, we introduce a so-called prox-QNEM algorithm, being a combination between an EM algorithm~\citep{dempster1977maximum}, an L-BFGS-B algorithm~\citep{zhu1997algorithm} and a proximal gradient descent algorithm~\citep{beck2009fast}. EM algorithm has already been used for multivariate data joint modeling (see~\citet{lin2002maximum} for instance), but here we face different and original problems: for each subject $i$, the latent variables are the pairs $(G_i, b_i)$ (not only the random effects); and then, we want to minimize the penalized objective $\ell_n^\text{pen}$ (not ``only'' the negative log-likelihood).

We first need to compute the negative completed log-likelihood (here scaled by $n^{-1}$), namely the negative joint distribution of $\cD_n$, $\textbf{\textit{b}} = (b_1, \ldots, b_n)$ and $\textbf{\textit{G}} = (G_1, \ldots, G_n)$.
It can be written
\begin{align*}
  \ell_n^\text{comp}(\theta) &= \ell_n^\text{comp}(\theta ; \cD_n, \textbf{\textit{b}}, \textbf{\textit{G}}) \\ 
  &= - n^{-1} \sum_{i=1}^n -\dfrac12 (r \log 2\pi + \log|D| + b_i^\top D^{-1}b_i) + \sum_{k=0}^{K-1} \ind{\{G_i=k\}} \Big[ \log \pi_{\xi_k}(x_i) \\ 
  & \quad + \delta_i \Big(\log \lambda_0(t_i) + \gamma_k^\top \Psi_i(t_i) \Big) - \int_0^{t_i} \lambda_0(s) \exp \big\{ \gamma_k^\top \Psi_i(s) \big\} \dd s \\
  & \quad + (y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \Big].
\end{align*}
Suppose that we are at step $w + 1$ of the algorithm, with current iterate denoted $\theta^{(w)}$.

\paragraph*{E-step.}

We need to compute the expected negative log-likelihood of the complete data conditional on the observed data and the current estimate of the parameters given by 
\begin{equation*}
  \cQ_n(\theta, \theta^{(w)}) = \E_{\theta^{(w)}}[\ell_n^\text{comp}(\theta) | \cD_n].
\end{equation*}
Given Assumption~\ref{assumption1}, the previous expression requires to compute expectations of the form
\[ \E_{\theta^{(w)}}[ g(b_i, G_i) | t_i, \delta_i, y_i] = \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \int_{\R^r} g(b_i, G_i) f(b_i | t_i, \delta_i, y_i ; \theta^{(w)}) \dd b_i \]
for different functions $g$, where we denote 
\begin{equation}
  \label{eq:pi_ik-def}
  \pi_{ik}^{\theta^{(w)}} = \P_{\theta^{(w)}}[G_i = k | t_i, \delta_i, y_i] 
\end{equation}
the posterior probability of the latent class membership using parameters $\theta^{(w)}$. 
Then, either $g(b_i, G_i) = \tilde g(b_i)$ (either $\tilde g_1: b_i \mapsto b_i$ or $\tilde g_2: b_i \mapsto b_i b_i^\top$), or $g(b_i, G_i) = g_k(G_i)$ with $g_k~:~G_i \mapsto \ind{\{G_i=k\}}$.

In the case $g(b_i, G_i) = \tilde g(b_i)$, one has
\begin{equation*}
  \E_{\theta^{(w)}}[ \tilde g(b_i) | t_i, \delta_i, y_i] = \int_{\R^r} \tilde g(b_i) f_{\theta^{(w)}}(b_i | t_i, \delta_i, y_i) \dd b_i = \dfrac{ \sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,\tilde g}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,1}^{\theta^{(w)}}}
\end{equation*}
where
\begin{equation}
  \label{eq:Lambda_ik-b_i}
  \Lambda_{ik,\tilde g}^{\theta^{(w)}} =  f_{\theta^{(w)}}(t_i, \delta_i | G_i = k) f_{\theta^{(w)}}(y_i | G_i = k) \int_{\R^r} \tilde g(b_i) f_{\theta^{(w)}}(b_i | y_i, G_i = k) \dd b_i,
\end{equation}
and in the Gaussian case used in practice one has
\begin{align*}
  y_i | G_i = k ; \theta^{(w)} &\sim \cN\big(U_i\beta_k^{(w)}, V_iD^{(w)}V_i^\top + \Sigma_i^{(w)}\big), \\
  b_i | y_i, G_i=k ; \theta^{(w)} &\sim \cN \big(O_{i,k}^{(w)}, W_i^{(w)}\big),  
\end{align*}
with $O_{i,k}^{(w)}=W_i^{(w)}V_i^T{\Sigma_i^{(w)}}^{-1}(y_i - U_i\beta_k^{(w)})$ and $W_i^{(w)} = (V_i^\top{\Sigma_i^{(w)}}^{-1}V_i + {D^{(w)}}^{-1})^{-1}$. Hence, what is interesting here is that~\eqref{eq:Lambda_ik-b_i} is actually closed-formed since 
\[ \E_{\theta^{(w)}}[ \tilde g_1(b_i) | y_i, G_i = k] = O_{i,k}^{(w)} \quad \text{and} \quad \E_{\theta^{(w)}}[ \tilde g_2(b_i) | y_i, G_i = k] = W_i^{(w)} + O_{i,k}^{(w)}{O_{i,k}^{(w)}}^\top\]
and then tractable analytically. Whereas in classical SREMs, some form of approximation must be used in practice to compute this quantity, like numerical integration techniques (for instance Gaussian quadrature) or Monte Carlo approximation.

>>>>

\section{section 1}
